---
title: "Law Estate Wines — Post-Waitlist Sales Cleaning"
subtitle: "Sales export, 2022-08 to 2025-10"
author: "Carly"
date: today
format:
  html:
    toc: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

## Overview

This cleans the post-waitlist sales export and writes `post_waitlist_sales_clean.csv` (and `.rds`),
**one row per line item**, using the same rules and column names as `wd_sales_clean`, so the two
can be stacked into one 2013–2025 table.

This export has fewer fields than the WD files, so some WD columns don't exist here:
no order time, taxes, order total, original/unit price, ship-to address, phone, birth date,
original order number, or club tier.

::: {.callout-note}
**Coverage:** starts 2022-08-01, where the WD data effectively ends (the last WD file has only
2 rows after 2022-07-31). **Order numbers restart at 1011 in October 2025**, so they collide with
2013 WD order numbers. When stacking, identify orders by `order_number` + dataset, not
`order_number` alone.
:::

::: {.callout-warning}
The output contains customer PII. Keep `post_waitlist_sales_clean.csv` / `.rds` in `.gitignore`.
:::

## Setup

```{r}
#| label: setup
library(readr)
library(dplyr)
library(stringr)
library(lubridate)
library(janitor)
library(here)
```

## Read

Reads every `Sales Post Waitlist ... .csv` in the project root (works whether the export is one
file or split into parts), all columns as text. Types are parsed explicitly below.

```{r}
#| label: read
paths <- list.files(here(), pattern = "^Sales Post Waitlist.*\\.csv$", full.names = TRUE)
stopifnot(length(paths) >= 1)
basename(paths)

raw <- paths |>
  lapply(\(p) read_csv(p, col_types = cols(.default = col_character()),
                       name_repair = "unique_quiet") |>
           mutate(source_file = basename(p))) |>
  bind_rows() |>
  clean_names() |>
  remove_empty("cols") |>                       # 3 blank trailing columns, empty Order Source Code
  rename(any_of(c(order_number = "r")))         # order-number header is truncated to "r"

dim(raw)
```

### De-duplicate

97 rows are exact duplicates, with whole orders repeated line for line (an export artifact).
`source_file` is ignored for the comparison.

```{r}
#| label: dedupe
raw <- raw[vctrs::vec_unique_loc(select(raw, -source_file)), ]
nrow(raw)
```

## Helpers

- **Money** comes as text: `"$1,234.00 "`, `"($88.00)"` for negatives, or a bare `"87"`.
- **Postal codes:** US zips padded to 5 digits and ZIP+4 truncated; Canadian and other
  international codes kept as-is.
- **States:** full names ("California") converted to codes; international regions
  ("Hong Kong", "Antwerp", "Dubai") get `bill_country = "Other"` and no state code.

```{r}
#| label: helpers
parse_money <- function(x) {
  x   <- str_trim(x)
  neg <- str_detect(x, "^\\(.*\\)$|^-")
  v   <- parse_number(x)
  if_else(coalesce(neg, FALSE), -abs(v), v)
}

us_codes     <- c(state.abb, "DC", "PR", "VI", "GU", "AS", "MP", "AA", "AE", "AP")
ca_provinces <- c("AB", "BC", "MB", "NB", "NL", "NS", "NT", "NU", "ON", "PE", "QC", "SK", "YT")

clean_state <- function(s) {
  s  <- str_squish(s)
  up <- str_to_upper(s)
  case_when(
    up %in% c(us_codes, ca_provinces)  ~ up,
    str_to_title(s) %in% state.name    ~ state.abb[match(str_to_title(s), state.name)],
    TRUE                               ~ NA_character_
  )
}

clean_postal <- function(z, country) {
  z <- str_to_upper(str_remove_all(z, "[^A-Za-z0-9]"))
  case_when(
    is.na(z) | z == ""              ~ NA_character_,
    country %in% c("CA", "Other")   ~ z,
    str_detect(z, "[A-Z]")          ~ z,
    nchar(z) == 4                   ~ str_pad(z, 5, pad = "0"),
    nchar(z) == 5                   ~ z,
    nchar(z) %in% 8:9               ~ str_sub(str_pad(z, 9, pad = "0"), 1, 5),
    TRUE                            ~ NA_character_
  )
}

clean_city <- function(x) str_to_title(str_squish(str_remove(x, ",\\s*$")))

wine_rx <- regex(
  "Beyond Category|First Tracks|The Nines|Nines|Cabernet Sauvignon|Aspire|Audacious|Beguiling|Intrepid|Sagacious|Prima|Cirque|Rose|Soph|\\bBEG\\b|\\bSAG\\b",
  ignore_case = TRUE
)
year_rx <- "(?<!\\d)20[0-2]\\d(?!\\d)"   # also matches "2021Beguiling" (no space)
```

## Row-level cleaning

- Drops 1,039 empty rows (no product, no quantity, $0).
- Parses dates, quantities, and money; negative quantity or price = refund.
- Fixes "Repeatcustomer" → "RepeatCustomer".
- Standardizes emails, cities, states, zips; adds `bill_country`.
- Sets zip **94111** to `NA` unless the city is San Francisco. It's the POS default for walk-ins
  (about 1,000 rows, mostly with no city or state).
- Sales associates: shared credit ("Addie/Renee", "Renee and Nicole D.") → "Multiple";
  first-name-only and spelling variants fixed.
- Flags house accounts (`is_internal`) and walk-in guest accounts (`is_guest_account`);
  neither gets a `customer_key`.

```{r}
#| label: clean-rows
lines <- raw |>
  filter(!is.na(product_name)) |>
  mutate(
    # ---- types ----
    order_number    = as.integer(order_number),
    customer_number = as.integer(parse_number(customer_number)),
    order_date      = mdy(completed_date),
    quantity_sold   = as.numeric(quantity_sold),
    across(c(ext_item_price, ext_item_shipping, item_discount), parse_money),
    # 110 $0 waitlist lines (Oct 2025) have blank shipping/discount
    across(c(ext_item_shipping, item_discount), ~ coalesce(.x, 0)),

    # ---- customer status ----
    contact_status = recode(contact_status, "Repeatcustomer" = "RepeatCustomer"),

    # ---- location ----
    bill_email      = na_if(str_to_lower(str_trim(bill_email)), ""),
    bill_city       = clean_city(bill_city),
    state_raw       = str_squish(bill_state_code),
    bill_state_code = clean_state(state_raw),
    bill_country    = case_when(
      bill_state_code %in% ca_provinces       ~ "CA",
      !is.na(bill_state_code)                 ~ "US",
      !is.na(state_raw) & state_raw != "0"    ~ "Other"
    ),
    bill_zip = clean_postal(bill_zip, bill_country),
    bill_zip = if_else(bill_zip == "94111" & coalesce(bill_city, "") != "San Francisco",
                       NA_character_, bill_zip),

    # ---- sales associates ----
    sales_associate = case_when(
      str_detect(sales_associate, "/|,| and ")  ~ "Multiple",
      sales_associate == "Kim"                  ~ "Kim Roth",
      sales_associate == "Kelli law"            ~ "Kelli Law",
      sales_associate == "KayDee Rickerd"       ~ "Kaye Dee Rickerd",
      sales_associate == "Pedro"                ~ "Pedro Vasquez",
      TRUE                                      ~ sales_associate
    ),

    # ---- house / guest accounts ----
    bill_full_name = str_to_lower(str_c(coalesce(bill_first_name, ""), " ",
                                        coalesce(bill_last_name, ""))),
    is_internal = coalesce(str_detect(bill_email, "@lawestatewines\\.com$"), FALSE) |
                  str_detect(bill_full_name, "bottle breakage|event marketing|employee samples|charity events") |
                  coalesce(bill_company == "Law Estate Wines", FALSE),
    is_guest_account = str_detect(bill_full_name, "^(guest customer|pos guest)$"),

    customer_key = if_else(
      is_internal | is_guest_account, NA_character_,
      coalesce(as.character(customer_number),
               if_else(!is.na(bill_email), str_c("email:", bill_email), NA_character_))
    ),

    # ---- flags ----
    is_refund = coalesce(quantity_sold < 0, FALSE) | coalesce(ext_item_price < 0, FALSE),
    is_comp   = coalesce(str_detect(sales_attribute, regex(
                  "sample|donation|replacement|corked|employee|gift/prize|poured at event|publication|team tasting",
                  ignore_case = TRUE)), FALSE)
  )
```

## Product standardization

Same categories as the WD cleaning, plus `adjustment` (tip adjustments, a wholesale partial
payment). New in this export: 3L bottles, "Magnum" written before the vintage, names without a
space ("2021Beguiling"), Cabernet Sauvignon as a regular wine, and many dated event tickets.

The first matching rule wins. Events come before tasting so "Winemaker Tasting … Ticket" counts
as an event, and shipping comes before wine sets so "Ship box- 12pack" counts as shipping.

```{r}
#| label: products
lines <- lines |>
  mutate(
    pn = str_to_lower(product_name),

    item_category = case_when(
      str_detect(pn, "cottage")                                               ~ "lodging",
      str_detect(pn, "tip adjustment|partial payment")                        ~ "adjustment",
      str_detect(pn, "ship box|cool pack|wooden ?box|topper|re-?route|shipping") ~ "shipping_packaging",
      str_detect(pn, "membership|waitlist|migrate [bw]lwc|library-|add on")   ~ "club_admin",
      str_detect(pn, "gift certificate|etching|engraving")                    ~ "merch",
      str_detect(pn, "insert|brochure|holiday card")                          ~ "promo_gift",
      str_detect(pn, "vertical|horizontal|bundle|\\dpack|package")            ~ "wine_set",
      str_detect(pn, "meal|lunch|food plate|cheese or chocolate")             ~ "food",
      str_detect(pn, paste0(
        "ticket|dinner|party|pick ?up|pup22|wine camp|new side of law|takeover|spotlight|",
        "inspiration|vine2vessel|shuttle|site fee|bld \\d|blade|gratuity|^ppf |",
        "(mon|tues|wednes|thurs|fri|satur|sun)day,"))                          ~ "event",
      str_detect(pn, paste0(
        "tasting|t&t|(new|return) (guest|member|industry)|member guest|",
        "^\\$\\d+ (new|return)|walking t/t|no show fee"))                      ~ "tasting",
      str_detect(pn, "^glass")                                                ~ "glass_pour",
      str_detect(pn, "gift|card")                                             ~ "promo_gift",
      str_detect(product_name, wine_rx) | str_detect(pn, year_rx)             ~ "wine",
      TRUE                                                                    ~ "other"
    ),

    tasting_waived = item_category == "tasting" & str_detect(pn, "waived"),

    wine_name = if_else(item_category %in% c("wine", "glass_pour", "wine_set"),
                        str_extract(product_name, wine_rx), NA_character_),
    wine_name = case_match(str_to_lower(wine_name),
                           c("nines", "the nines") ~ "The Nines",
                           "beg"                   ~ "Beguiling",
                           "sag"                   ~ "Sagacious",
                           .default = str_to_title(wine_name)),

    vintage = if_else(item_category %in% c("wine", "glass_pour"),
                      as.integer(str_extract(pn, year_rx)), NA_integer_),

    bottle_format = if_else(item_category == "wine",
                            case_when(str_detect(pn, "\\b3l\\b")          ~ "3L",
                                      str_detect(pn, "magnum|1\\.5l")     ~ "1.5L",
                                      TRUE                                ~ "750ml"),
                            NA_character_),

    # canonical name: "2021 Audacious", "2021 Audacious Magnum", "2020 Intrepid 3L"
    product_std = case_when(
      item_category == "wine" ~ str_squish(str_c(
        coalesce(as.character(vintage), ""), " ", wine_name,
        case_when(bottle_format == "1.5L" ~ " Magnum",
                  bottle_format == "3L"   ~ " 3L",
                  TRUE                    ~ ""))),
      TRUE ~ str_squish(str_remove_all(product_name,
               regex("\\(?\\s*-?\\s*sold out!*\\s*-?\\s*\\)?", ignore_case = TRUE)))
    )
  ) |>
  select(-pn)
```

## Final table

Same column names and order as `wd_sales_clean` for every field this export has.

Dropped: `completed_date` (→ `order_date`), `product_type` and `department` (→ `item_category`),
pickup/ship dates and tracking number (logistics), `order_notes` (free text), plus the helper
columns.

```{r}
#| label: final
sales_clean <- lines |>
  select(
    # order
    source_file, order_number, order_type, order_date,
    sales_associate, sales_attribute, pickup,
    # customer
    customer_key, customer_number, contact_status, is_internal, is_guest_account,
    bill_first_name, bill_last_name, bill_company, bill_email,
    bill_city, bill_state_code, bill_zip, bill_country,
    # product
    product_sku, product_name, product_std, item_category,
    wine_name, vintage, bottle_format, tasting_waived,
    # money
    quantity_sold, item_discount, ext_item_price, ext_item_shipping,
    # flags
    is_refund, is_comp
  )
```

## Sanity checks

### Size and date range

```{r}
#| label: check-size
cat("Lines:", nrow(sales_clean),
    " Orders:", n_distinct(sales_clean$order_number),
    " Customers:", n_distinct(sales_clean$customer_key, na.rm = TRUE), "\n")
range(sales_clean$order_date)
count(sales_clean, year = year(order_date))
```

### Line items by category

```{r}
#| label: check-category
count(sales_clean, item_category, sort = TRUE)
```

### Wine names

Look for one wine split across two `product_std` values.

```{r}
#| label: check-wines
count(filter(sales_clean, item_category == "wine"), product_std, sort = TRUE)
```

### Unmatched products

```{r}
#| label: check-other
distinct(filter(sales_clean, item_category == "other"), product_name)
```

### Parsing problems

These should all be 0.

```{r}
#| label: check-parsing
sales_clean |>
  summarise(
    bad_date     = sum(is.na(order_date)),
    bad_qty      = sum(is.na(quantity_sold)),
    bad_price    = sum(is.na(ext_item_price)),
    bad_shipping = sum(is.na(ext_item_shipping))
  )
```

### Flags

```{r}
#| label: check-flags
sales_clean |>
  summarise(refund = sum(is_refund), comp = sum(is_comp),
            internal = sum(is_internal), guest_account = sum(is_guest_account))
count(sales_clean, bill_country)
```

## Write output

```{r}
#| label: write
# .rds keeps column types (zips stay text with leading zeros, dates stay dates)
saveRDS(sales_clean, here("post_waitlist_sales_clean.rds"))
write_csv(sales_clean, here("post_waitlist_sales_clean.csv"))
```
