---
title: "Law Estate Wines — Wine Club Member Segmentation"
author: "Carly"
date: today
format:
  html:
    toc: true
    df-print: paged
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

## Overview

Groups wine club members by **purchasing behavior** using k-means, then checks the result with
hierarchical clustering (Ward's method).

**Data:** the two cleaned sales datasets (`wd_sales_clean.rds`, 2013–2022, and
`post_waitlist_sales_clean.rds`, 2022–2025), stacked and aggregated to **one row per member**.
The member files (active/cancelled) are better used to *describe* the clusters afterward, for
example "which segment cancels most", than as clustering inputs, because cancellation is an
outcome, not a behavior.

**Who's included:**

- Anyone with a `ClubMember` contact status or a `ClubOrder` on record.
- At least one order since 2022-08-01, so the segments describe the current membership.
- At least 2 orders and some wine purchased, so per-year rates are meaningful.
- House accounts and walk-in guest accounts are excluded.

## Setup

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(lubridate)
library(cluster)   # silhouette(); ships with R
library(here)

set.seed(518)

# ---- parameters: change these to redefine the population ----
min_last_order <- as.Date("2022-08-01")   # member must have ordered since this date
min_orders     <- 2
k_grid         <- 2:8
```

## Load and stack sales

Order numbers restart in Oct 2025 in the post-waitlist data, so orders are identified by
dataset + order number (`order_uid`).

```{r}
#| label: load
wd <- readRDS(here("wd_sales_clean.rds"))            |> mutate(dataset = "WD")
pw <- readRDS(here("post_waitlist_sales_clean.rds")) |> mutate(dataset = "PW")

common <- intersect(names(wd), names(pw))
sales  <- bind_rows(wd[common], pw[common]) |>
  mutate(order_uid = str_c(dataset, "-", order_number))

as_of <- max(sales$order_date, na.rm = TRUE)
cat("Lines:", nrow(sales), " Date range:", format(range(sales$order_date)), "\n")

# customer numbers should carry across both exports (same WineDirect account IDs)
cat("Customers in both datasets:",
    length(intersect(wd$customer_key, pw$customer_key)), "\n")
```

## Identify members

```{r}
#| label: members
lines <- sales |>
  filter(!is.na(customer_key), !is_internal, !is_guest_account)

member_keys <- lines |>
  group_by(customer_key) |>
  summarise(is_member = any(contact_status == "ClubMember", na.rm = TRUE) |
                        any(order_type == "ClubOrder", na.rm = TRUE)) |>
  filter(is_member) |>
  pull(customer_key)

lines <- filter(lines, customer_key %in% member_keys)
length(member_keys)
```

## Build member features

Order-level fields (order type, pickup) repeat on every line, so they're collapsed to one row
per order first. Orders made up entirely of refunds are dropped.

Rates are **per active year**: first order to last order, with a 1-year floor so new members
aren't inflated. Recency is kept as its own feature, so a lapsed member shows up as high
recency rather than low frequency.

```{r}
#| label: features
orders <- lines |>
  group_by(customer_key, order_uid) |>
  summarise(
    order_date  = first(order_date),
    order_type  = first(order_type),
    pickup      = first(pickup),
    has_tasting = any(item_category == "tasting"),
    has_event   = any(item_category == "event"),
    all_refund  = all(is_refund),
    .groups = "drop"
  ) |>
  filter(!all_refund)

# wine purchases, in 750ml-equivalent bottles (magnum = 2, 3L = 4)
wine <- lines |>
  filter(item_category == "wine", !is_comp) |>
  mutate(btl_eq = quantity_sold * case_when(bottle_format == "1.5L" ~ 2,
                                            bottle_format == "3L"   ~ 4,
                                            TRUE                    ~ 1)) |>
  group_by(customer_key) |>
  summarise(
    bottles      = sum(btl_eq),
    wine_spend   = sum(ext_item_price),
    wine_variety = n_distinct(wine_name[quantity_sold > 0], na.rm = TRUE),
    .groups = "drop"
  )

spend <- lines |>
  filter(!is_comp) |>
  group_by(customer_key) |>
  summarise(net_spend = sum(ext_item_price), .groups = "drop")

# profile-only attributes (not clustered on)
geo <- lines |>
  arrange(desc(order_date)) |>
  group_by(customer_key) |>
  summarise(state = first(na.omit(bill_state_code)),
            zip   = first(na.omit(bill_zip)), .groups = "drop") |>
  mutate(in_ca    = coalesce(state == "CA", FALSE),
         local_slo = coalesce(str_starts(zip, "934"), FALSE))   # SLO County zips

feat <- orders |>
  group_by(customer_key) |>
  summarise(
    first_order      = min(order_date),
    last_order       = max(order_date),
    n_orders         = n(),
    club_order_share = mean(order_type == "ClubOrder"),
    online_share     = mean(order_type == "Website"),
    pickup_share     = mean(pickup == "Yes", na.rm = TRUE),
    tasting_visits   = sum(has_tasting),
    events           = sum(has_event),
    .groups = "drop"
  ) |>
  left_join(wine,  by = "customer_key") |>
  left_join(spend, by = "customer_key") |>
  left_join(geo,   by = "customer_key") |>
  mutate(
    active_yrs     = pmax(as.numeric(last_order - first_order) / 365.25, 1),
    tenure_yrs     = as.numeric(as_of - first_order) / 365.25,
    recency_days   = as.numeric(as_of - last_order),
    orders_per_yr  = n_orders / active_yrs,
    bottles_per_yr = coalesce(bottles, 0) / active_yrs,
    spend_per_yr   = net_spend / active_yrs,
    avg_btl_price  = if_else(coalesce(bottles, 0) > 0, wine_spend / bottles, NA_real_),
    tasting_per_yr = tasting_visits / active_yrs,
    events_per_yr  = events / active_yrs,
    pickup_share   = coalesce(pickup_share, 0)
  ) |>
  filter(last_order >= min_last_order,
         n_orders   >= min_orders,
         coalesce(bottles, 0) > 0,
         avg_btl_price > 0)

nrow(feat)
```

## Prepare the clustering matrix

K-means uses Euclidean distance, so features must be on comparable scales and not dominated
by a few extreme buyers:

1. `log1p()` on the skewed count and rate features.
2. Winsorize at the 1st/99th percentile, so a handful of collectors don't get their own cluster.
3. Standardize (mean 0, SD 1).

`spend_per_yr` is left out because it's roughly bottles × price, which are both already in.

```{r}
#| label: prep
log_vars   <- c("recency_days", "tenure_yrs", "orders_per_yr", "bottles_per_yr",
                "avg_btl_price", "tasting_per_yr", "events_per_yr", "wine_variety")
share_vars <- c("club_order_share", "pickup_share", "online_share")
cluster_vars <- c(log_vars, share_vars)

winsor <- function(x, p = c(0.01, 0.99)) {
  q <- quantile(x, p, na.rm = TRUE)
  pmin(pmax(x, q[1]), q[2])
}

X <- feat |>
  select(all_of(cluster_vars)) |>
  mutate(across(all_of(log_vars), log1p),
         across(everything(), winsor)) |>
  scale()

stopifnot(!anyNA(X))
dim(X)
```

### Feature correlations

Pairs above about |0.8| are nearly redundant. Consider dropping one from `cluster_vars`.

```{r}
#| label: corr
#| fig-height: 6
cor(X) |>
  as.data.frame() |>
  tibble::rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  ggplot(aes(var1, var2, fill = r)) +
  geom_tile() +
  geom_text(aes(label = round(r, 2)), size = 2.5) +
  scale_fill_gradient2(limits = c(-1, 1)) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Choosing k

Two views: the elbow in within-cluster sum of squares, and average silhouette width
(higher is better). Silhouette is computed on a 3,000-member sample for speed.

```{r}
#| label: choose-k
km_fits <- lapply(k_grid, \(k) kmeans(X, centers = k, nstart = 50, iter.max = 100))

sil_idx <- sample(nrow(X), min(3000, nrow(X)))
d_sil   <- dist(X[sil_idx, ])

k_eval <- tibble(
  k   = k_grid,
  wss = sapply(km_fits, `[[`, "tot.withinss"),
  silhouette = sapply(km_fits, \(f)
    mean(silhouette(f$cluster[sil_idx], d_sil)[, "sil_width"]))
)
k_eval

k_eval |>
  pivot_longer(-k) |>
  ggplot(aes(k, value)) +
  geom_line() + geom_point() +
  facet_wrap(~ name, scales = "free_y") +
  scale_x_continuous(breaks = k_grid) +
  theme_minimal()
```

Pick `k` from the plots and from how interpretable the profiles below are. The default is the
k with the highest silhouette. Override it if a neighboring k gives more useful groups.

```{r}
#| label: k-final
k_final <- k_eval$k[which.max(k_eval$silhouette)]
# k_final <- 4   # <- override here
k_final

km <- km_fits[[which(k_grid == k_final)]]
feat$cluster <- factor(km$cluster)
table(feat$cluster)
```

## Check with hierarchical clustering

Ward's method on the same matrix. If most members land in matching groups under both methods,
the segments are real structure and not an artifact of k-means' random starts.

```{r}
#| label: hclust
hc <- hclust(dist(X), method = "ward.D2")
feat$hc_cluster <- factor(cutree(hc, k = k_final))

table(kmeans = feat$cluster, hierarchical = feat$hc_cluster)

# share of members whose hierarchical group is the best match for their k-means group
tab <- table(feat$cluster, feat$hc_cluster)
cat("Agreement:", round(sum(apply(tab, 1, max)) / sum(tab), 3), "\n")
```

## Visualize

The first two principal components of the clustering matrix, colored by k-means cluster.

```{r}
#| label: pca
pca <- prcomp(X)
pve <- round(100 * summary(pca)$importance[2, 1:2], 1)

as_tibble(pca$x[, 1:2]) |>
  mutate(cluster = feat$cluster) |>
  ggplot(aes(PC1, PC2, color = cluster)) +
  geom_point(alpha = 0.4, size = 1) +
  labs(x = str_glue("PC1 ({pve[1]}%)"), y = str_glue("PC2 ({pve[2]}%)")) +
  theme_minimal()
```

## Cluster profiles

### What distinguishes each cluster

Cluster centers in standardized units. Positive means above the average member, and ±1
means one SD from average.

```{r}
#| label: profile-heatmap
#| fig-height: 6
km$centers |>
  as.data.frame() |>
  mutate(cluster = factor(row_number())) |>
  pivot_longer(-cluster, names_to = "feature", values_to = "z") |>
  ggplot(aes(cluster, feature, fill = z)) +
  geom_tile() +
  geom_text(aes(label = round(z, 1)), size = 3) +
  scale_fill_gradient2() +
  labs(x = "Cluster", y = NULL) +
  theme_minimal()
```

### In real units

Medians per cluster, plus the profile-only attributes (spend, location).

```{r}
#| label: profile-table
feat |>
  group_by(cluster) |>
  summarise(
    members            = n(),
    pct_of_spend       = sum(net_spend) / sum(feat$net_spend),
    recency_days       = median(recency_days),
    tenure_yrs         = median(tenure_yrs),
    orders_per_yr      = median(orders_per_yr),
    bottles_per_yr     = median(bottles_per_yr),
    spend_per_yr       = median(spend_per_yr),
    avg_btl_price      = median(avg_btl_price),
    wine_variety       = median(wine_variety),
    club_order_share   = median(club_order_share),
    pickup_share       = median(pickup_share),
    online_share       = median(online_share),
    tasting_per_yr     = median(tasting_per_yr),
    events_per_yr      = median(events_per_yr),
    pct_in_ca          = mean(in_ca),
    pct_local_slo      = mean(local_slo)
  ) |>
  mutate(across(where(is.double), \(x) round(x, 2)))
```

After reading the profiles, give each cluster a name here, for example "Local regulars",
"Club-only shippers", "Lapsed", "Collectors":

```{r}
#| label: names
cluster_names <- c(
  "1" = "Cluster 1",
  "2" = "Cluster 2",
  "3" = "Cluster 3",
  "4" = "Cluster 4",
  "5" = "Cluster 5",
  "6" = "Cluster 6",
  "7" = "Cluster 7",
  "8" = "Cluster 8"
)[seq_len(k_final)]

feat <- mutate(feat, segment = cluster_names[as.character(cluster)])
count(feat, segment)
```

## Save

```{r}
#| label: save
member_segments <- feat |>
  select(customer_key, segment, cluster, hc_cluster, all_of(cluster_vars),
         spend_per_yr, net_spend, n_orders, first_order, last_order, in_ca, local_slo)

saveRDS(member_segments, here("member_segments.rds"))
readr::write_csv(member_segments, here("member_segments.csv"))
```
